Computing solution space properties of combinatorial optimization problems via generic tensor networks

### 2022-05-12 paper.tex
change codes

\begin{abstract}
%We introduce a method that uses generic programming tensor networks to compute various properties of a class of hard graph problems that can be decribed by local constraints.
%We introduce a tensor network algorithm %a method that combines generic programming and tensor network techniques 
We introduce a unified framework to compute the solution space properties of a broad class of combinatorial optimization problems.
%including the independent set problem, the maximum cut problem, the vertex coloring problem, the clique problem, the dominating set problem, the set packing problem, the set covering problem, and the satisfiability problem.
%Let the independent set problem be an example, we show how to compute, for example, the maximum several independent set sizes, the counting of independent sets of a give size, the enumeration of independent sets of a given size.
These 
%solution space 
properties include finding one of the optimum solutions, counting the number of solutions of a given size, and enumeration and sampling of solutions of a given size. Using 
%We use 
the independent set problem as an example, we show how all these solution space properties can be computed in the unified approach of generic tensor networks.
%By making use of generic programming, our algorithms are very simple to implement and one can in addition directly utilize recent advances in tensor network contraction techniques such as near-optimal contraction order finding and slicing to achieve high performance.
%By using generic programming techniques, the same framework can be used to compute all of these properties. 
%Our algorithm utilizes recent advances in tensor network contraction techniques to achieve high performance, including methods to quickly find a near-optimal contraction order and slicing.
%The algorithmic complexity of this approach is two to the treewidth of the line graph of the tensor network graph.
We demonstrate the versatility of this computational tool by applying it to several  examples, including computing the entropy constant for hardcore lattice gases, studying the overlap gap properties, and analyzing the performance of quantum and classical algorithms for finding maximum independent sets.
% including the calculations of the entropy constant for some hardcore lattice gases on 2D square lattices and the study of the overlap gap property on three regular graphs.
%We show how our method helps to understand these hard problems on systems with hundreds of vertices.
\end{abstract}

@@

% REQUIRED
% 14N07  	Secant varieties, tensor rank, varieties of sums of powers
\begin{AMS}
  05C31, 14N07
\end{AMS}

@@

\begin{figure} 
    \centering
    \includegraphics[width=\textwidth, trim={0cm 0cm 0cm 0cm}, clip]{figures/fig1.pdf}
    \caption{Benchmark results for computing different solution space properties of independent sets of random three-regular graphs with different tensor element types.
    The time in these plots only includes tensor network contraction, without taking into account the contraction order finding and just-in-time compilation time.
    Legends are properties, algebra, and devices that we used in the computation; one can find the corresponding computed solution space property in \Tbl{tbl:generictypes}.
    (a) time and space complexity versus the number of vertices for the benchmarked graphs.
    (b) The computation time for calculating the MIS size and for counting the number of all independent sets (ISs), the number of MISs,
        the number of independent sets having size $\alpha(G)$ and $\alpha(G)-1$, and finding $100$ largest set sizes.
    (c) The computation time for calculating the independence polynomials with different approaches.
    (d) The computation time for configuration enumeration, including single MIS configuration, the enumeration of all independent set configurations, all MIS configurations, all independent sets,
    and all independent set configurations having size $\alpha(G)$ and $\alpha(G)-1$.
    }
    \label{fig:benchmark}
\end{figure}

@@

\begin{figure}[t] 
    \centering
    \includegraphics[width=\textwidth, trim={0cm 1cm 0cm 1cm}, clip]{figures/lattices.pdf}
    \caption{The types of graphs used in the case study in section \ref{sec:entropy}.
    The lattice dimensions are $L\times L$. (a) square lattice graphs. (b) square lattice graphs with a filling factor $p=0.8$.
    (c) King's graphs. (d) King's graphs with a filling factor $p=0.8$.}
    \label{fig:lattices}
\end{figure}

\begin{figure}[t] 
    \centering
    \includegraphics[width=\textwidth, trim={0cm 0cm 0cm 0cm}, clip]{figures/fig5.pdf}
    \caption{Mean entropy for lattice gases on graphs defined in \Fig{fig:lattices}.
    We sampled $1000$ instances for $p=0.8$ lattices and the error bar is too small to be visible.
    The horizontal black dashed lines are for $\lim_{L\rightarrow \infty} F(L,L)^{1/L^2}$ for the corresponding non-disordered square lattice and King's graphs.
    }
    \label{fig:hardsquare}
\end{figure}

@@

    \includegraphics[width=\textwidth, trim={0.0cm 1cm 0.0cm 0cm}, clip]{figures/fig6.pdf}

@@

%\subsection{Visualizing experimental quantum algorithm outputs for solving Maximum Independent Set}
\subsection{Analyzing quantum and classical algorithms for Maximum Independent Set}
\begin{figure} 
    \centering
    \includegraphics[width=.65\textwidth, trim={0cm 0cm 0cm 0cm}, clip]{figures/exp_configurations.pdf}
    \caption{Visualization of experimental outputs of a quantum algorithm for solving the maximum independent set problem~\cite{Ebadi2022}.
    Each vertex represents an independent set, and each edge represents a pair of independent sets that differ by a swap operation or a vertex addition.}
    \label{fig:exp_configuratoins}
\end{figure}
In a recent work, 
the ability to enumerate configurations and compute independence polynomials was critical in understanding the performance of quantum optimization algorithms for the MIS problem on a Rydberg atom quantum computer~\cite{Ebadi2022}. This work focused on 
%, the authors found that
exploring King's graphs with $0.8$ filling.   The hardest instances for classical simulated annealing could be accurately predicted from the independence polynomial, which gave information about the density of local minima at different independent set sizes.
On the hardest graph instances for simulated annealing studied in the experiment, a high density of local minima were found at independent set sizes of $\alpha(G)-1,$ which the algorithm became trapped in instead of finding the optimal solution of size $\alpha(G)$.
%Initially, it was suspected that simulated annealing trapped certain clusters of these local minima in the solution space. However, 
By enumerating the configurations using techniques described in the present work, we found that %n found that there was no clustering in the solution space as in Sec.~\ref{sec:overlap-gap}, and instead 
simulated annealing randomly explores the independent sets of size $\alpha(G)-1$ until an optimum solution is found.
Therefore, the large ratio of local to global minima  prevents simulated annealing from efficiently finding an MIS.
%and studying the solution space properties for independent sets of size $\alpha(G)$ and $\alpha(G)-1$, we gained detailed insights to performance of both the quantum algorithm and simulated annealing, and could determine if these algorithm were likely to return independent sets from certain clusters of local or global minima in the solution space. Instead, we found that simulated annealing becomes trapped in independent sets of size $\alpha(G)-1$ uniformly at random, which is related to observation that there are no disconnected clusters in the solution space for King's graphs in Sec.~\ref{sec:overlap-gap}.

@@

\begin{figure} 
    \centering
    \includegraphics[width=\textwidth, trim={0cm 0cm 0cm 0cm}, clip]{figures/fig2.pdf}
    \caption{Benchmarks of computing different solution space properties of the maximal independent sets (ISs) problem on random three regular graphs at different sizes.
    (a) time and space complexity of tensor network contraction. 
    (b) The wall clock time for counting and enumeration of maximal ISs. % The missing data is due to the memory limit.
    }
    \label{fig:benchmark-maximal}
\end{figure}

@@

\section{Bounding the MIS enumeration space}\label{sec:bounding}
\begin{figure}
    \centering
    \includegraphics[width=0.9\textwidth, trim={0cm 0cm 0cm 0cm}, clip]{figures/masktree.pdf}
    \caption{Bounded enumeration of maximum independent sets. Here, a circle is a tensor, an arrow specifies the execution direction of a function, $\overline A$ is the Boolean mask for $A$ and $\circ$ is the Hadamard (element-wise) multiplication. (a) is the forward pass with tropical algebra (\Eq{eq:tropical}) for computing $\alpha(G)$.
     (b) is the backward pass for computing Boolean gradient masks.
     (c) is the masked tensor network contraction with tropical algebra combined with sets (\Eq{eq:countingtropicalset}) for enumerating configurations.}
     \label{fig:bounding}
\end{figure}

@@

\centerline{\includegraphics[width=0.8\textwidth, trim={0cm 1cm 0cm 1cm}, clip]{figures/fig7.pdf}}

@@

\lstinputlisting[breaklines]{../democode/demo.jl}

>>>>

\begin{abstract}
We introduce a unified framework to compute the solution space properties of a broad class of combinatorial optimization problems.
These properties include finding one of the optimum solutions, counting the number of solutions of a given size, and enumeration and sampling of solutions of a given size. Using the independent set problem as an example, we show how all these solution space properties can be computed in the unified approach of generic tensor networks.
We demonstrate the versatility of this computational tool by applying it to several  examples, including computing the entropy constant for hardcore lattice gases, studying the overlap gap properties, and analyzing the performance of quantum and classical algorithms for finding maximum independent sets.
\end{abstract}

@@

% REQUIRED
% 14N07  	Secant varieties, tensor rank, varieties of sums of powers
\begin{AMS}
  15A69, 05C31, 14N10
\end{AMS}

@@

\begin{figure} 
    \centering
    \includegraphics[width=\textwidth, trim={0cm 0cm 0cm 0cm}, clip]{fig1.pdf}
    \caption{Benchmark results for computing different solution space properties of independent sets of random three-regular graphs with different tensor element types.
    The time in these plots only includes tensor network contraction, without taking into account the contraction order finding and just-in-time compilation time.
    Legends are properties, algebra, and devices that we used in the computation; one can find the corresponding computed solution space property in \Tbl{tbl:generictypes}.
    (a) time and space complexity versus the number of vertices for the benchmarked graphs.
    (b) The computation time for calculating the MIS size and for counting the number of all independent sets (ISs), the number of MISs,
        the number of independent sets having size $\alpha(G)$ and $\alpha(G)-1$, and finding $100$ largest set sizes.
    (c) The computation time for calculating the independence polynomials with different approaches.
    (d) The computation time for configuration enumeration, including single MIS configuration, the enumeration of all independent set configurations, all MIS configurations, all independent sets,
    and all independent set configurations having size $\alpha(G)$ and $\alpha(G)-1$.
    }
    \label{fig:benchmark}
\end{figure}

@@

\begin{figure}[t] 
    \centering
    \includegraphics[width=\textwidth, trim={0cm 1cm 0cm 1cm}, clip]{lattices.pdf}
    \caption{The types of graphs used in the case study in section \ref{sec:entropy}.
    The lattice dimensions are $L\times L$. (a) square lattice graphs. (b) square lattice graphs with a filling factor $p=0.8$.
    (c) King's graphs. (d) King's graphs with a filling factor $p=0.8$.}
    \label{fig:lattices}
\end{figure}

\begin{figure}[t] 
    \centering
    \includegraphics[width=\textwidth, trim={0cm 0cm 0cm 0cm}, clip]{fig5.pdf}
    \caption{Mean entropy for lattice gases on graphs defined in \Fig{fig:lattices}.
    We sampled $1000$ instances for $p=0.8$ lattices and the error bar is too small to be visible.
    The horizontal black dashed lines are for $\lim_{L\rightarrow \infty} F(L,L)^{1/L^2}$ for the corresponding non-disordered square lattice and King's graphs.
    }
    \label{fig:hardsquare}
\end{figure}

@@

    \includegraphics[width=\textwidth, trim={0.0cm 1cm 0.0cm 0cm}, clip]{fig6.pdf}

@@

%\subsection{Visualizing experimental quantum algorithm outputs for solving Maximum Independent Set}
\subsection{Analyzing quantum and classical algorithms for Maximum Independent Set}
\begin{figure} 
    \centering
    \includegraphics[width=.65\textwidth, trim={0cm 0cm 0cm 0cm}, clip]{exp_configurations.pdf}
    \caption{Visualization of experimental outputs of a quantum algorithm for solving the maximum independent set problem~\cite{Ebadi2022}.
    Each vertex represents an independent set, and each edge represents a pair of independent sets that differ by a swap operation or a vertex addition.}
    \label{fig:exp_configuratoins}
\end{figure}
In a recent work, 
the ability to enumerate configurations and compute independence polynomials was critical in understanding the performance of quantum optimization algorithms for the MIS problem on a Rydberg atom quantum computer~\cite{Ebadi2022}. This work focused on 
%, the authors found that
exploring King's graphs with $0.8$ filling.   The hardest instances for classical simulated annealing could be accurately predicted from the independence polynomial, which gave information about the density of local minima at different independent set sizes.
On the hardest graph instances for simulated annealing studied in the experiment, a high density of local minima were found at independent set sizes of $\alpha(G)-1,$ which the algorithm became trapped in instead of finding the optimal solution of size $\alpha(G)$.
%Initially, it was suspected that simulated annealing trapped certain clusters of these local minima in the solution space. However, 
By enumerating the configurations using techniques described in the present work, we found that %n found that there was no clustering in the solution space as in Sec.~\ref{sec:overlap-gap}, and instead 
simulated annealing randomly explores the independent sets of size $\alpha(G)-1$ until an optimum solution is found.
Therefore, the large ratio of local to global minima  prevents simulated annealing from efficiently finding an MIS.
%and studying the solution space properties for independent sets of size $\alpha(G)$ and $\alpha(G)-1$, we gained detailed insights to performance of both the quantum algorithm and simulated annealing, and could determine if these algorithm were likely to return independent sets from certain clusters of local or global minima in the solution space. Instead, we found that simulated annealing becomes trapped in independent sets of size $\alpha(G)-1$ uniformly at random, which is related to observation that there are no disconnected clusters in the solution space for King's graphs in Sec.~\ref{sec:overlap-gap}.

@@

\begin{figure} 
    \centering
    \includegraphics[width=\textwidth, trim={0cm 0cm 0cm 0cm}, clip]{fig2.pdf}
    \caption{Benchmarks of computing different solution space properties of the maximal independent sets (ISs) problem on random three regular graphs at different sizes.
    (a) time and space complexity of tensor network contraction. 
    (b) The wall clock time for counting and enumeration of maximal ISs. % The missing data is due to the memory limit.
    }
    \label{fig:benchmark-maximal}
\end{figure}

@@

\section{Bounding the MIS enumeration space}\label{sec:bounding}
\begin{figure}
    \centering
    \includegraphics[width=0.9\textwidth, trim={0cm 0cm 0cm 0cm}, clip]{masktree.pdf}
    \caption{Bounded enumeration of maximum independent sets. Here, a circle is a tensor, an arrow specifies the execution direction of a function, $\overline A$ is the Boolean mask for $A$ and $\circ$ is the Hadamard (element-wise) multiplication. (a) is the forward pass with tropical algebra (\Eq{eq:tropical}) for computing $\alpha(G)$.
     (b) is the backward pass for computing Boolean gradient masks.
     (c) is the masked tensor network contraction with tropical algebra combined with sets (\Eq{eq:countingtropicalset}) for enumerating configurations.}
     \label{fig:bounding}
\end{figure}

@@

\centerline{\includegraphics[width=0.8\textwidth, trim={0cm 1cm 0cm 1cm}, clip]{fig7.pdf}}

@@

\lstinputlisting[breaklines]{demo.jl}
